Experiences Gamifying Developer Adoption\\ of Practices and Tools

### 2013-10-23 paper_id51.tex
Submitted version at 4:30 Tuesday

As software development practices evolve, toolsmiths face the continuous challenge of communicating new practices and tools to the community.   We tested an idea with industrial software developers that adding game-like feedback to the development environment would improve adoption of tools and practices for code navigation.  We present results from a pre-study survey of 130 developers' opinions on gamification and motivation, data from a pilot by an intact team of six developers of a usage monitoring tool with game features tied to code navigation practices, and feedback collected in post-pilot interviews.  
Most developers expressed interest in gamification in our survey, though some have strong negative opinions.  Pilot results show that two of the six pilot developers adjusted their practices when presented with competitive game elements.
%We found that a software engineering game with competitive elements influences some developers towards changing their practices.
%We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
%
% In addition to training, discussing, and presenting software engineering practices and tools, we demonstrate a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).  
%
%We determine through a survey of 130 developers the things we can do to motivate them to adopt 
%Blaze monitors developer practice and tool use over the long term and reports the lasting change a tool deployment or communication effort makes in everyday habits. 
%
% Embedding Blaze in the developer's work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.   Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.  
% With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.  
%
%Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.  We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools. The Blaze tool was well received by most participants, however, navigation practices were not significantly affected by the Blaze feedback mechanism.  We conclude that more proactive interventions are necessary to encourage developers to change their code navigation habits.
\end{abstract}

% A category with the (minimum) three required fields
\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\terms{Human Factors, Management}

\keywords{empirical, training, learning, usability}

@@

Like Sascha, software developers who maintain code face challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Toolsmiths respond to these challenges by creating new tools for assisting developers through difficult activities like code navigation.  However, as new practices and tools become available, developers may be motivated to start using them unless they interact with a peer developer who promotes them\cite{wbsnipes:Hill2011Peer}.    Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.  Thus we need to both communicate with and motivate developers like Sascha to start using new tools by showing them how they benefit from using the new tools.

@@

Would software developers writing serious software be motivated by this gamification concept? To answer this question, this work contributes an assessment of developers' response to game-like elements tied to their use of new tools and practices.  We assess 130 developers' response and relevant concerns about the idea of gamification through a web-based survey.  We piloted Blaze, our tool that introduces game elements focused on improving navigation practices into the developer's IDE.  Developers' reactions to gamification were mostly positive with a few detractors, and results of piloting show that game elements influence some developers but have no effect on others' use of tools and practices.

@@

The two areas of related work we discuss are research studies using other monitoring methods and studies applying motivation and gamification to other areas.

@@

Murphy-Hill, Parnin, and Black \cite{V:MurphyHill2012How} use the Mylyn Monitor to explore whether or not developers use the automated refactoring tools present in Eclipse.  They look for specific refactoring commands in Eclipse and determine the amount of time developers use tools versus hand refactoring the code.  For this study we focus on code navigation as a high-value area for improving developer efficiency.

The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.  Studies conducted using PROM include a series of studies on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.  Studies using the PROM tool correlate the time spent editing with code metrics obtained trough source code analysis.  Applying these data to specific research questions such as whether refactoring improves productivity \cite{Moser2008Case} shows the utility of combining automated effort data with code metric data.  Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect how the user was navigating through the code in addition to knowing how long the editor is open for a particular file.

Murphy-Hill et al. study a large usage history data set and apply several different algorithms to accurately suggest IDE commands to novice users.\cite{MurphyHill2012Improving} Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history. Novice users were better served by recommendation algorithms that did not require a long history, where more expert users benefited from more sophisticated algorithms that included more history. The ``Most Widely Used'' algorithm recommends commands based on the collective usage profile of the team and performed nearly as well as the more sophisticated algorithms. The paper presents the methodology for making recommendations and evaluate their accuracy. Our work goes takes a further step evaluating how well tools can influence developers to use a specific recommended practice.

@@

When discussing gamification, folks raise a concern about replacing an activity like software development, which is intrinsically motivated, with extrinsic motivation provided by points and achievements.  To discover what is behind this concern, several studies are useful to understand what should be done.

Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{Beecham2008Motivation}.  They claim most common motivator found in cited references is ''the work'' motivates us. The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making. The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills and tackle new challenges.

Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group choosing the behavior that meets the expectations and values of the group that is significant to them\cite{wbsnipes:MaehrCulture}.  A key factor in raising achievement motivation is establishing a social group where the person is motivated to belong and excel.  Maehr states that since achievement behaviors can be triggered by circumstance, they are present in most people regardless of background.

Hamari et al. analyze achievements patterns in game systems and provide guidance on defining successful achievement systems.  They recommend  achievement awards be setup as a secondary form of scoring so players have multiple objectives available for a particular game.  Players should feel like they earned achievements in the game, they should not be given for trivial tasks.

@@

Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}. Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.  They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.  Dubois and Tamburrelli demonstrate this guidance using a study of students working on a class project.  Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.  The subject group can see each others' rankings while the control group only sees their personal rankings.  Results though inconclusive, show competition increases the desired results.  In our study we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.

@@

Using a question from the Saatchi and Saatchi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB ``How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace''?  We segmented results by country in Figure \ref{fig:gamification} and found gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.  The responses indicate developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.  For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.

Comments from participants reflected a more interesting controversial reaction than these measurements might indicate.  A developer from India said ``Gamification is a juvenile way of getting programmers to do something that you want.''  A different respondent expressed a concern that ``Gamification of best practices can have nasty side effects.''  Another participant said ``there is already high a level of competition between software development teams..'' indicating they are under considerable pressured to get the work done.  Another developer reacted with a statement that identifies the kind of intrinsic motivation we hope for ``I am using practices and tools because I want to do my job well',' indicating that gamification was unnecessary to motivate people to improve.

On the positive side, a developer from Poland said ``In a previous work place we had a 'game' between developers - scoring broken/unbroken builds and new/passed/failed tests. It was helping to improve quality and It worked well. Recommended especially for junior developers. ''  Other comments indicated competition or points would be acceptable ``as long as it doesn't interrupt or slow down my work. ''

Because we nearly identically reused their original question, we compare the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.  Our results land on both sides of the 75\% mark resulting from the Saatchi and Saatchi study.

@@

Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.  The question provided a graduated scale for the scope of who data were shared with and divided the questions between whether sharing was anonymous or not. Figure \ref{fig:comfortwithsharing}  shows over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members. 
Their comfort level decreases as the scope of who the information is shared with increases.  People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data were not kept anonymous.  
The first conclusion from these questions on sharing data is people are very willing to share information that could help their team. Second, sharing within the company is acceptable if we take care to make the data anonymous.

@@

We asked developers what would motivate them to use tools and practices and gave them four choices to rank.  The top motivator for using tools and practices is coworker recommendation with 75\% of respondents in Figure \ref{fig:toolandpracticemotivators}   ranking in 1st or 2nd place.  Of several comments, a developer in Sweden captured the following scene, ``[The] most motivating [thing] is when a new practice / tool is discussed during a continuous improvement meeting or even coffee break, where one co-developer tells about the benefits and there is a joint discussion on where it (tool or practice) could be best applied.''

@@

Just behind team competition is management mandate where developers would be asked to follow specific practices by the organization.  The lowest was having badges posted on their social network profile where 32\% ranked this in 1st or 2nd place.   Badges and management mandate received the most negative written comments in the survey.  A developer from the United States captured the lower opinion of badges that several participants expressed ``I am motivated because ABB allows me to choose the tools that best work for me. A badge will not really do anything for me.''  Badges also ranked low in motivations for taking more software engineering training.

We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.

@@

Developers also ranked the desirability of awards for different actions.  Overall 68\% of developers ranked receiving an award for consistently performing quality practices as either 1st or 2nd place.  Thus, focusing awards on good practices would motivate developers to do them consistently.

 To summarize, the high-level conclusions from each key question are:
\begin{itemize}[itemsep=0mm]
\item 95\% of developers responding would try tools and practices suggested by an automated recommendation system 
\item Developers are motivated by collaboration and team goals more than mandates and individual awards
\item Awards for performing good quality development practices ranked highest
\item Over 90\% of respondents approved of sharing data with their team members
\end{itemize}
Thus, this pre-study survey provided a strong positive indicator for going forward with a software engineering tool focused on sharing developer data and rewarding developers for using good practices and tools.

@@

We defined the first criteria for selecting a practice based on prior work  applying gamification to other domains outside software development. 
In their paper on game design patterns, Hamari et al. issue a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework}.  Thus we avoided selecting practices like bug fixing or task completion because they are simply requirements of the job.

@@

To evaluate developers using recommended tools in Visual Studio, we categorized them by their tool name.  In the pilot we wanted developers to learn about and adopt the Sando search tool.  Sando is part of the recommended developer tool suite for ABB and supports structured navigation using code search \cite{Shepherd2012Sando}.

Gamification practices consider the feedback of achievements and points as critical to a successful outcome.  The design of achievement awards should gradually encourage the participant to reach higher and higher levels of success in the activity.  Achievements must feel earned to the participant so they recognize the effort to get the award was significant and feel satisfied  \cite{wbsnipes:Hamari2011Framework}.  Hamari and Eranti describe a general relationship between achievements in games and regular game play      In games achievements are typically a parallel scoring system to the main game play.  They make the game more engaging providing multiple ways to increase your score and multiple challenges in one interface.

Using these guidelines we assigned points to structured navigation commands which received one point each while unstructured navigation commands received zero points.    We gave additional emphasis to using the Sando search tool by assigning it ten points.  To create levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we applied an exponential curve based on points scored.  Users initially ``level up'' after  working a day or two; the next level may be achieved after a week or even a month following an exponential curve.  We designed the points and levels so above average developers can complete all levels during the study period.

By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition between teammates.    The leaderboard allows developers to see their own score in relation to the top five people on the team.  However, the identifier of the other participants was generated as three capital letters assigned based on their position on the leaderboard.  Not knowing who of your colleagues was ahead of you could be  reduce motivation, but this avoids one potential source of conflict in an intact team.

@@

Part of the research objective was to determine whether we can achieve adoption of a practice with alternative interventions.  As previously discussed, we sought to provide Blaze to influence developers towards using more structured navigation practices.  We constructed the interventions during the pilot to roll out in stages each week they use the tool.  This allowed us to establish a baseline and determine the changes as the pilot continued.

In the initial stage, participants were informed of the purpose of the study what to expect from the Blaze tool.  The participants installed the tool in their Visual Studio environment.  The tool did not present any window for them to see and simply collected data in the background.

After a week's data collection marking the end of the first stage, Blaze started to automatically pop-up a window containing a button for a web page with information.  The page included a link to download Sando  and information about using Sando in a video demo.  The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.  The About Blaze page contained details of the point scoring system and background on Blaze's purpose.  The usefulness of the communication site is rated as part of our post-pilot  survey.

In the third stage developers received instant feedback from Blaze on the points they accumulated during the development session.  The tool's appearance at this stage is shown in Figure \ref{fig:blazeWindow}.  The Info button provides a level indicator via colored contents and activates a chart display windows when clicked.  Points shows the points accumulated for the developer's history.  Level is the level of the game achieved based on an exponentially growing curve.   The developer could click on the Leaderboard tab to see how their scores compared with the others in the pilot.  This final configuration remained as the display until the end of the pilot.

@@

As mentioned in Figure \ref{fig:gamification}, India is one country where as per our survey, the acceptance to gamification is the highest. Hence, we conducted this study with an intact team of 6 developers working in an R\&D facility in India on a large industrial software system. We approached a team who develops and maintains an industrial software product. Initially, we met the whole team along with the management staff to explain the research project and the Blaze tool we developed. We  explained the aim of the study, what data is collected, and how the data will be processed and managed.

It was important to gain the confidence of the participants that their data were kept confidential so their activities would be close as possible to real-world. if the participants felt like they are being tracked and the data would identify them to their peers or managers, then they may refuse to participate or deviate from their normal development style.

We  assured the team about the confidentiality of the data and the management reassured the data collected by the research group will not be used for any other purpose other than the study.  Developers were asked to volunteer for the pilot by their management, however, managers did not know who participated in the study. Their name and data were kept confidential to ensure no one apart from the research team knows he/she is participating in the pilot. Mechanisms in our data collection to identify individuals were under the control of each developer, they could chose to share or keep confidential the association of a generated unique id with researchers or each other.

Management wished to have no knowledge of who was participating and took great care to remain detached from the pilot proceedings.  Data at this level of detail has not been collected in ABB development organizations.  Researchers were careful to create the system where developers controlled their identification with the data and fortunately, management bought into the necessary anonymity of the participants.

@@

The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers' navigation practices and knowledge.  The survey questions were segmented into portions that evaluate prior and post knowledge and practices for navigation, key steps in the pilot itself, the influence of game elements on their navigation practices, typical tasks and demographic information.

To find out whether Blaze had an effect on knowledge of the developers regarding structured navigation commands, we asked them to rate their knowledge of structured navigation commands prior to the pilot on a 1-10 scale to understand their baseline.  Then we asked them to contrast their knowledge after the pilot to assess what they learned from the pilot.

The other key question is did Blaze have a perceived effect on their navigation practices.  Here too we asked them to rate how much they used structured navigation prior to the pilot using a 4 choice rating of not at all, some, many times, and as much as possible.  Then we asked whether the used structured navigation commands about the same amount, more,  or a lot more during the pilot.  These two questions assessed their perceptions of structured navigation command use.  We compared the answers with the usage data collected from Blaze.

The final assessment questions tested whether using Blaze influenced their attention to navigation commands by asking how much specific game elements influenced their practices.  The scale started with not at all, a little, some, to a lot.

The other questions included whether they reviewed the training materials on structured navigation commands, installed the Sando code search tool, and what types of development tasks they would consider using structured navigation for in general.  Another demographic question is how well they know the code they work in rated from ``I work in code that I wrote'', ``I maintain code that I am familiar with'', to ``I recently learned the code''

@@

We asked developers to rate their knowledge of the code they maintain.  Four of the six developers reported they maintain code that they wrote while the two most experienced developers maintain code that they recently learned.  One developer reported during our individual analysis feedback that working in code they know well affects their navigation practices leading them to use more unstructured navigation.

On the assessment of structured navigation knowledge, half of the participants replied they had very good prior knowledge about structured navigation and they use structured navigation as much as possible.  Other half of the participants have some understanding on structured navigation use it some of the time. Even developers whose navigation ratio  was lower than average, used structured navigation occasionally as much as their colleagues.  While analyzing the points they scored with Blaze tool, we observed the participants who rated their prior knowledge of structured navigation highly also scored higher in the pilot.

@@

We selected navigation as the subject of the pilot game because it was identified by Robillard et al.\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.    In this analysis we ask whether we are focusing on the right area of developers' activity in Visual Studio?

>>>>

As software development practices evolve, toolsmiths face the continuous challenge of getting developers to adopt new practices and tools.   We tested an idea with industrial software developers that adding game-like feedback to the development environment would improve adoption of tools and practices for code navigation.  We present results from a pre-study survey of 130 developers' opinions on gamification and motivation, usage data from a pilot with an intact team of six developers of a game on code navigation practices, and feedback collected in post-pilot interviews.  Our pre-study survey showed that most developers were interested in gamification, though some have strong negative opinions.  Pilot results show that two of the six pilot developers adjusted their practices when presented with competitive game elements.
%We found that a software engineering game with competitive elements influences some developers towards changing their practices.
%We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
%
% In addition to training, discussing, and presenting software engineering practices and tools, we demonstrate a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).  
%
%We determine through a survey of 130 developers the things we can do to motivate them to adopt 
%Blaze monitors developer practice and tool use over the long term and reports the lasting change a tool deployment or communication effort makes in everyday habits. 
%
% Embedding Blaze in the developer's work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.   Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.  
% With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.  
%
%Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.  We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools. The Blaze tool was well received by most participants, however, navigation practices were not significantly affected by the Blaze feedback mechanism.  We conclude that more proactive interventions are necessary to encourage developers to change their code navigation habits.
\end{abstract}

\category{D.2.9}{Software}{Software Engineering} {Management} [Productivity]

\terms{Management, Economics, Human computer interaction}

\keywords{Development, Productivity, Code Navigation}

@@

Like Sascha, software developers who maintain code face challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Toolsmiths respond to these challenges by creating new tools for assisting developers through difficult activities like code navigation.  However, as new practices and tools become available, developers may be motivated to start using them unless they interact with a peer developer who promotes them\cite{wbsnipes:Hill2011Peer}.    Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.  Thus, we need to both communicate with and motivate developers like Sascha to start using new tools.

@@

Would software developers writing serious software be motivated by this gamification concept? To answer this question, this work contributes an assessment of developers' response to game-like elements tied to their use of new tools and practices.  We assess 130 developers' response and relevant concerns about the idea of gamification through a web-based survey.  We piloted Blaze, our tool that introduces game elements focused on improving navigation practices into the developer's IDE.  Developers' reactions to gamification were mostly positive with a few detractors, and results of piloting show that game elements influence some developers, but have no effect on others' use of tools and practices.

@@

The two areas of related work we discuss are research studies using other monitoring methods and studies applying gamification to other areas.

@@

Murphy-Hill, Parnin, and Black \cite{V:MurphyHill2012How} use the Mylyn Monitor to explore whether or not developers use the automated refactoring tools present in Eclipse.  They look for specific refactoring commands in Eclipse and determine the amount of time developers use tools versus hand refactoring the code.  For this study, we focus on code navigation as a high-value area for improving developer efficiency.

The Pro Metrics (PROM) tool provides a framework for collecting data for further analysis from tools used by developers.\cite{Coman2009Casestudy}  It provides a flexible data model and a plug-in architecture to facilitate collection from different data sources.  Studies conducted using PROM include a series of studies on trends in time spent Pair Programming \cite{Coman2008Investigating}, benefits of refactoring on productivity \cite{Moser2008Case}, impact of refactoring on re-usability \cite{Moser2006Does}, and prediction of effort \cite{Abrahamsson2007Effort}.  Studies using the PROM tool correlate the time spent editing with code metrics obtained through source code analysis.  Applying these data to specific research questions ,such as whether refactoring improves productivity \cite{Moser2008Case}, shows the utility of combining automated effort data with code metric data.  Taking a different focus on techniques, our study used more detailed events captured from the IDE  to detect commands the uses to navigating through the code in addition to knowing duration of edit sessions for code modules.

Murphy-Hill et al. study a large usage history data set and apply several different algorithms to accurately suggest IDE commands to novice users\cite{MurphyHill2012Improving}.  Algorithms based on command history performed well particularly when synchronized chronologically with the recipient's usage history. Novice users were better served by recommendation algorithms that did not require a long history, while more expert users benefited from more sophisticated algorithms that included more history. The ``Most Widely Used'' algorithm recommends commands based on the collective usage profile of the team and performed nearly as well as the more sophisticated algorithms. The paper presents the methodology for making recommendations and evaluating their accuracy. Our work takes a further step evaluating how well tools can influence developers to use a specific recommended commands.

@@

When discussing gamification, folks raise a concern about replacing an activity like software development, which is intrinsically motivated, with extrinsic motivation provided by points and achievements.  To discover what is behind this concern, we reviewed several studies on motivation and gamification.

Beecham and colleagues provide a thorough analysis of existing literature on studies of motivation in Software Engineering \cite{Beecham2008Motivation}.  They claim that the most common motivator found in cited references is ''the work''. The list of de-motivators is also useful with common job satisfaction items like stress and inequity in recognition, plus poor quality software (low accomplishment) and lack of influence in decision making. The paper also lists characteristics of people in the professions including need for independence (autonomy), desire to learn new skills and tackle new challenges.

Maehr proposes an affirmative theory whereby  individuals achieve as a member of a social group, choosing the behavior that meets the expectations and values of the group that is significant to them\cite{wbsnipes:MaehrCulture}.  A key factor in raising achievement motivation is establishing a social group where the person is motivated to belong and excel.  Maehr states that since achievement behaviors can be triggered by circumstance, they are present in most people regardless of background.

Hamari et al. analyze achievements patterns in game systems and provide guidance on defining successful achievement systems.  They recommend  achievement awards be setup as a secondary form of scoring so players have multiple objectives available for a particular game.  Players should feel like they earned achievements in the game; they should not be given for trivial tasks.

@@

Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}. Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.  They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.  Dubois and Tamburrelli demonstrate this guidance using a study of students working on a class project.  Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.  The subject group can see each others' rankings while the control group only sees their personal rankings.  Results, though inconclusive, show competition increases the desired results.  In our study, we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.

@@

Using a question from the Saatchi and Saatchi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB ``How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace''?  We segmented results by country in Figure \ref{fig:gamification} and found gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.  The responses indicate 75-88\% of developers in India, Poland, and Finland were at least somewhat interested.  For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.

Comments from participants reflected a more interesting controversial reaction than these measurements might otherwise indicate.  A developer from India said ``Gamification is a juvenile way of getting programmers to do something that you want.''  A different respondent expressed a concern that ``Gamification of best practices can have nasty side effects.''  Another participant said ``there is already high a level of competition between software development teams..'' indicating they are under considerable pressured to get the work done.  Another developer reacted with a statement that identifies the kind of intrinsic motivation we hope for ``I am using practices and tools because I want to do my job well',' indicating that gamification was unnecessary to motivate people to improve.

On the positive side, a developer from Poland said, ``In a previous work place we had a 'game' between developers - scoring broken/unbroken builds and new/passed/failed tests. It was helping to improve quality and it worked well. [It is] recommended especially for junior developers. ''  Other comments indicated competition or points are acceptable ``as long as it doesn't interrupt or slow down my work. ''

Because we nearly identically reused their original question, we compare the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.  Our results land on both sides of the 75\% mark reported by the Saatchi and Saatchi study.

@@

Another question was whether sharing detailed usage information with colleagues and the company as a whole would be a concern for developers.  The question provided a graduated scale for the scope of who data was shared with and divided the questions between whether sharing was anonymous or not. Figure \ref{fig:comfortwithsharing}  shows over 90\% of respondents are comfortable with sharing either anonymous or non-anonymous information with their team members. 
Their comfort level decreases as the scope of who the information is shared with increases.  People were less comfortable sharing with anyone (the category for people outside ABB) particularly if the data was not kept anonymous.  
The first conclusion from these questions on sharing data is that people are very willing to share information that could help their team. Second, sharing within the company is acceptable if we take care to make the data anonymous.

@@

We asked developers what would motivate them to use tools and practices and gave them four choices to rank.  In Figure \ref{fig:toolandpracticemotivators}, we see that 75\% of respondents ranked coworker recommendation in 1st or 2nd place.  Of several comments, a developer in Sweden captured the following scene, ``[The] most motivating [thing] is when a new practice / tool is discussed during a continuous improvement meeting or even coffee break, where one co-developer tells about the benefits and there is a joint discussion on where it (tool or practice) could be best applied.''

@@

Just behind team competition is management mandate where developers would be asked to follow specific practices by the organization.  The lowest was having badges posted on their social network profile where 32\% ranked this in 1st or 2nd place.   Badges and management mandate received the most negative written comments in the survey.  A developer from the United States captured the lower opinion of badges that several participants expressed ``I am motivated because ABB allows me to choose the tools that best work for me. A badge will not really do anything for me.''  We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.

@@

In another question, developers ranked several actions that they would like to receive awards for.  Overall 68\% of developers ranked receiving an award for consistently performing quality practices as either 1st or 2nd place.  Thus, providing awards to developers for using good practices would motivate developers to do them consistently.

 To summarize, the high-level conclusions from each key question are:
\begin{itemize}[itemsep=0mm]
\item 95\% of developers responding would try tools and practices suggested by an automated recommendation system 
\item Developers are motivated by collaboration and team goals more than mandates and individual awards
\item Awards for good quality development practices ranked highest
\item Over 90\% of respondents approved of sharing data with their team members
\end{itemize}
Thus, this pre-study survey provided a positive indicator for going forward with a software engineering tool focused on sharing developer data and rewarding developers for using good practices and tools.

@@

We defined the first criteria for selecting a practice based on prior work  applying gamification to other domains outside software development. 
In their paper on game design patterns, Hamari et al. issued a warning that assigning achievements to required tasks reduces intrinsic motivation because player autonomy is reduced by the achievements \cite{wbsnipes:Hamari2011Framework}.  Thus we avoided selecting practices like bug fixing or task completion because they are simply requirements of the job.

@@

To evaluate developers use of recommended tools in Visual Studio, we categorized them by their tool name.  As part of the pilot, we wanted developers to learn about and adopt the Sando search tool.  Sando is part of the recommended developer tool suite for ABB and supports structured navigation using code search \cite{Shepherd2012Sando}.

Gamification practices consider the feedback of achievements and points as critical to a successful outcome.  The design of achievement awards should gradually encourage the participant to reach higher and higher levels of success in the activity.  Achievements must feel earned to the participant so they recognize the effort to get the award was significant and feel satisfied  \cite{wbsnipes:Hamari2011Framework}.  Hamari and Eranti describe a general relationship between achievements in games and regular game play.     Achievements in games typically create a parallel scoring system to the main game play.  They make the game more engaging by providing multiple ways to increase your score and multiple challenges in one interface.

Considering these guidelines, we assigned one point for each use of structured navigation commands while unstructured navigation commands received zero points.    We gave additional emphasis to using the Sando search tool by assigning ten points for each use.  To create levels following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we applied an exponential curve based on points scored.  Users initially ``level up'' after  working a day or two; the next level may require a week's work to achieve as the difficulty increases.  We designed the points and levels so above average developers can pass all levels during the study period.

By selecting an intact team for the pilot, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition between teammates.    The leaderboard allows developers to see their own score in relation to the top five people on the team.  However, the other participants' identifiers are auto-generated as three capital letters assigned based on their position on the leaderboard.  Not knowing who of your colleagues was ahead of you could reduce motivation, but this avoids one potential source of conflict in an intact team.

@@

Part of the research objective was to determine whether we can drive adoption of a practice through gamification.  As previously discussed, designed the game in Blaze to influence developers towards using more structured navigation practices.  We constructed the interventions during the pilot to roll out in stages each week they use the tool.  This allowed us to establish a baseline and determine the changes as the pilot continued.

In the initial stage, participants were informed of the purpose of the study, and what to expect from the Blaze tool.  The participants installed the tool in their Visual Studio environment.  During this stage, Blaze did not present a window for them to see and simply collected data in the background.

After a week's data collection marking the end of the first stage, Blaze started to automatically pop-up a window containing a button for a web page with information.  The page included a link to download Sando  and information about using Sando in a video demo.  The sub-page on navigation contained a page of tutorial on the built-in structured navigation commands in Visual Studio.  The ``About Blaze'' page contained details of the point scoring system and background on Blaze's purpose.  The usefulness of the communication site is rated as part of our post-pilot  survey.

In the third stage, developers received instant feedback from Blaze on the points they accumulated during the development session.  The tool's appearance at this stage is shown in Figure \ref{fig:blazeWindow}.  The ``Info'' button provides a level indicator via colored contents and activates a chart display windows when clicked.  Points shows the points accumulated for the developer's history.  Level is the level of the game achieved based on an exponentially growing curve.   The developer could click on the Leaderboard tab to see how their scores compared with the others in the pilot.  This final configuration remained as the display until the end of the pilot.

@@

As mentioned in Figure \ref{fig:gamification} from our survey, the acceptance of gamification is highest in India. Hence, we conducted this study with an intact team of 6 developers working in an R\&D facility in India on a large industrial software system. Initially, we met the whole team along with the management staff to explain the research project and the Blaze tool we developed. We  explained the aim of the study, what data is collected, and how the data would be processed and managed.

It was important to gain the confidence of the participants that their data were kept confidential so their activities would be as close as possible to real-world. If the participants felt like they are being tracked and the data would identify them to their peers or managers, then they may refuse to participate or deviate from their normal development style.

We  assured the team about the confidentiality of the data, and the management reassured them that the data collected by the research group will not be used for any other purpose other than the study.  Developers were asked to volunteer for the pilot by their management, however, managers did not know who participated in the study. Their names and data were kept confidential to ensure that no one apart from the research team knows he/she is participating in the pilot.  To distinguish individuals in the data each user is assigned a generated unique ID.  Developers could chose to share the unique ID with us or keep it confidential.  All were willing to share their unique ID with us during the follow-up feedback sessions.

Management wished to have no knowledge of who was participating and took great care to remain detached from the pilot proceedings.  Data at this level of detail has not been collected in ABB development organizations.  Researchers were careful to create the system where developers controlled whether we could associate their identity with their data, and fortunately management bought into the necessity for participant anonymity.

@@

The goal for the pilot follow-up survey was to collect feedback on the tool and the effect it had on the developers' navigation practices and knowledge.  The survey questions were segmented into portions that evaluate prior and post knowledge of navigation practices, key steps in the pilot itself, the influence of game elements on their navigation practices, typical tasks and demographic information.

To find out whether Blaze increased developers'  knowledge of structured navigation commands, we asked them to rate their knowledge of structured navigation commands prior to the pilot on a 1-10 scale.  Then we asked them to contrast their knowledge after the pilot to assess what they learned from the pilot.

The other key question is did Blaze have a perceived effect on their navigation practices.  Here too we asked them to rate how much they used structured navigation prior to the pilot using a 4 choice rating of ``not at all'', ``some'', ``many times'', and ``as much as possible''.  Then we asked whether they used structured navigation commands ``about the same amount'', ``more'',  or ``a lot more'' during the pilot.  These two questions assessed their perceptions of structured navigation command use.  We compared the answers with the usage data collected from Blaze.

The final assessment questions tested whether using Blaze influenced their attention to navigation commands by asking how much specific game elements influenced their practices.  The responses available were ``not at all'', ``a little'', ``some'', and ``a lot''.

The other questions included whether they reviewed the training materials on structured navigation commands, installed the Sando code search tool, and what types of development tasks they would consider using structured navigation for in general.  Another demographic question asked how well they know the code they work in rated from ``I work in code that I wrote'', ``I maintain code that I am familiar with'', to ``I recently learned the code''.

@@

We asked developers to rate their knowledge of the code they maintain.  Four of the six developers reported they maintain code that they wrote, while the two most experienced developers maintain code that they recently learned.  One developer reported during our individual feedback sessions that working in code they know well affects their navigation practices leading them to use more unstructured navigation.

On the assessment of prior structured navigation knowledge, half of the participants replied they had very good prior knowledge about structured navigation and they use structured navigation ``as much as possible''.  The other half of the participants rated their prior knowledge of structured navigation as good and reported that they use it ``some'' of the time. Of the three developers who rated their prior knowledge and use of structured navigation higher, two of them had the highest point scores of the group.

@@

We selected navigation as the subject of the pilot game because it was identified by Robillard et al.\cite{wbsnipes:Robillard2004How} as correlated with developers who effectively completed a maintenance task.    In this analysis, we ask whether we are focusing on the right area of developers' activity in Visual Studio?
